New lower bounds for Schur and weak Schur numbers

### 2021-05-07 article/weakschur.tex
jolis tableaux :'(
anglais + erreurs

Let \((p, q) \in \mathbb{N}^2\) such that there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free subsets. Let \(a \in \mathbb{N}\) with \(a > q\)
and let's try to build a colouring of \([\![1, ap + q]\!]\) into \(n + k\) weakly sum-free subsets. Let
\(l = a - b - 1\), \(r \in [\![1,q]\!]\) and \(w = a - l - r - 1 = b - r\).

@@

\underline{\textbf{\(\mathcal{T}\) block}} \\
We color this block using the weakly sum-free colouring of \([\![1,q]\!]\) with colors \(1, ..., n\).

\underline{\textbf{\(\mathcal{R}\) block}} \\
In this block,  we use the colors \(n + 1, ..., n + k\). We colour an integer \(x\) according to its line number (written \(\lambda(x)\)).
For every \(x \in \mathcal{R}\), we colour \(x\) with \(n + c\) where \(c\) is the colour of \(\lambda(x)\) in the sum-free colouring of  \([\![1,p]\!]\).
Let \((x, y) \in \mathcal{R}^2\). The cases are twofold.

\begin{itemize}
	\item \underline{\(\lambda(x+y) = \lambda(x) + \lambda(y)\)} \\
	In this case, we use the sum-free property of the colouring of \([\![1,p]\!]\) (in block \(\mathcal{C}\), we only use colours \(1, ..., n\)).

	\item \underline{\(\lambda(x+y) \neq \lambda(x) + \lambda(y)\)} \\
	In this case, we do not have information about the colour of \(\lambda(x+y)\). Thereby, we want to have \(x+y \in \mathcal{C}\).
	A simple solution is to limit the horizontal movement, that is if the sum changes line, not to move too far so that it stays in \(\mathcal{C}\).
	There, the maximal displacement to the left (resp. to the right) is \(2l\) (resp. \(2r\)). Not crossing entirely \(\mathcal{C}\) by going to the left
	is then expressed as \(-2l > -a + r\). Likewise, not going to far to the right is expressed as \(2r < a - l\). It can then be written as
	 \(\text{max}(l, r) \leqslant w\).
\end{itemize}

\underline{\textbf{\(\mathcal{C}\) block}} \\
In this block,  we use colors \(1, ..., n\). We colour an integer \(x\) according to its column number, written \(\tilde{\pi}(x)\). It is linked to the
projection on the first line, written \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\). A simple solution is to colour \(x\) with the same colour
as \(\tilde{\pi}(x)\) in the weakly sum-free colouring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colours \(1, ..., n\) are sum-free.

@@

\begin{definition}
Let \( (p,n,b) \in (\mathbb{N}^*)^3\), Let \((A_1,...,A_n)\) a partition of  \([\![1, p]\!]\).
This partition is said to be a b-weakly-sum-free template (b-WSF-template) of \(n\) colors and lenght \(p\) when:
\begin{itemize}
\item \(\forall i \in [\![1, n]\!], \quad A_i\) is weakly-sum-free
\item \(\forall i \in [\![1, n]\!], \quad A_i\backslash [\![1, b]\!]\) is sum-free
\item For \(A_n\) (the special subset): \[\forall (x,y) \in A_n^2, \,
x+y>b+2(p-b) \implies x+y-2(p-b)\notin A_n\]
\item For the others subsets: \\ \[\forall i \in [\![0,n-1]\!], \, \forall(x,y) \in A_i^2, \,
x+y>p \implies \pi(x+y) \notin A_i
\]
\end{itemize}
\end{definition}

@@

\begin{definition}
Let \( (k,b) \in (\mathbb{N}^*)^2\). If there exist \(p\) such that exists a partition of \([\![1, p]\!]\) into \(k\)
subsets which is a \(b\)-WSF-template of \(k\) colors and lenght \(p\), we note:
\\\\\(\WS_b^+=-b+\max \{p\in \mathbb{N}^*\)/ there exists a partition of \([\![1, p]\!]\) into \(k\) subsets which is a
\(b\)-WSF-template of \(k\) colors and lenght \(p\) \}
\\\\
If this \(p\) does not exist, we set \(\WS_b^+= 0\)
\end{definition}

\begin{definition}
Let \( n \in \mathbb{N}^*\), we define \(\WS^+(n)=\max_{b\in \mathbb{N}^*} \{\WS_b^+(n)\}\)
\end{definition}

\begin{proposition}
	Let \(n \in [\![2, +\infty]\!]\), we have :
	\[
	\frac{3}{2} \WS (n-1)+1 \leqslant \WS^+(n) \leqslant \WS (n)
	\]
\end{proposition}

\begin{proof}[\textsc{Proof.}]
The lower bound comes from the analogous of Abott and Hanson's construction for weak Schur numbers.
The upper bound comes from the fact that a WSF-template of length \(p\) with \(n\) colors is also a partition of
\([\![1, p]\!]\) into \(n\) sum-free subsets.
\end{proof}

@@

\begin{theorem}
Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \(k\)
sum-free subsets of \([\![1,p]\!]\) and a partition of \(n\) subsets \((A_1,....,A_n)\) of \([\![1, q]\!]\) which is a
\(b\)-WSF-template of \(n\) colors and lenght \(q\),
 then there exists a partition of \([\![1, b+p \times (q-b)]\!]\) into \((k+n-1)\) weakly sum-free subsets.
\end{theorem}

In particular, if we choose \(p = S(k)\) and \(q = \WS^+(n)\) in the last theorem, the next corollary follows.

@@

\begin{remark}
In the SF-template construction for Schur numbers, the additive constant comes from the fact that the special color does
not necessarily appear right at the begining of the repeating pattern. Likewise, \(b_{max}\) can actually be replaced by \\
\[\max \left\{ b + \min (A_{n+1} \backslash [\![1, b ]\!]) - 1~|~ \WS_b^+(n+1) = \WS^+(n+1) \right\}\]
\end{remark}

\begin{proof}[\textsc{Proof.}]
Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\), let \(a=q-b\). \\
We denote by \(g\) the colouring associated to the partition of \([\![1,q]\!]\) and \(h\) the
one associated to the partition of \([\![1,p]\!]\).
\[ g : [\![1,q]\!] \longrightarrow [\![1,n]\!] \text{ and } (A_{g^{-1}(1)},...A_{g^{-1}(q)})\text{ is a \(b\)-WSF-template.}
\]
\[h : [\![1,p]\!] \longrightarrow [\![1,k]\!] \text{ and } \forall (x,y) \in [\![1,q]\!]^2 \text{, } h(x) = h(y)
\Longrightarrow h(x+y) \neq h(x)
\]

@@

\begin{proposition}
Let \((b, k, p) \in \mathbb{N}^*)^3\) and let \(f\) be a colouring associated to a \(b\)-WSF-template of length \(p\) with \(k\) colors. Let
\(c \in \mathbb{N}\) (\(c = \min (A_{k+1} \backslash [\![1, b ]\!]) - 1\) works) and assume there there exists a colouring \(g\) of
\([\![b + 1, b + c]\!]\) with \(k\) colors such that:

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n))\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

\par These modifications work as follows. One may notice that the first row (excluding the "tail") has constraints that other rows
do not have because of the tail, especially if the special color appears in the tail as well. Thus allowing to have a colouring on the
first row different from the colouring of the other rows would weaken the constraints. Acutally, one may even go further by
noticing that on the one hand the first (ordered) color of the sum-free partition used for the extension procedure has more
more constraints than the other colors of the sum-free partition since the first row is of this color and is more constrained than
the other rows, but that on the other hand it has more degrees of freedom than the other colors of the sum-free partition since in
the sum-free partition there cannot be two consecutive numbers of this color. As a result, it removes  some constraints imposed
 by the first row on the other rows.

\par
To sum up, one can look for a generalised WSF-template that uses a special colouring for the tail and the first row, a
colouring dedicated to the rows whose number is not 1 but is in the first color in the sum-free partition, a colouring for all
the other rows and a special colouring for the last numbers (as previously explained for the improvement of the additive
constant of WSF-templates).

@@

\begin{theorem}
	Let \((n,k), (p,q) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template of \(k+1\) colors and lenght \(p\),
	and \(WSF\)-template of \(n\) color and lenght \(q\), then there exists \(WSF\)-template of \((n+k)\) and lenght \(pq\).
\end{theorem}

@@

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ \WS^+(n+k) \geqslant S^+(k+1) \WS^+(n) \]
\end{corollary}

@@

The first two inequalities were found by Rowley, they are detailed in \cite{RowleyWS}. The third inequality is optimal and was found with a SAT solver.
It uses the first sophistication explained in the previous subsection in order to add the last number in the first color.
As for the fourth inequality, it was obtained by combining an optimal SF-template of length 33 with a WSF-template of length 4.
The best template we could get with a computer search gives the inequality \(\WS(n+4) \geqslant 127 S(n) + 68\).
It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of
5 colors which start with a good \(\WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}.
This could be the suject of a future work. Further details about the encoding as a SAT problem can be found in \cite{Heule2017}.

@@

\begin{center}
\begin{tabular}{|*{5}{c|}}
    \hline
	n & 8 & 9 & 10 & 11 \\
	\hline
	\(4S(n-1) + 2 \) & 6\,722 & 21\,146 & \cellcolor{yellow} 71\,214 & \cellcolor{yellow} 243\,794\\
	\hline
	\(13S(n-2) + 8 \) & \cellcolor{yellow} 6\,976 & 21\,848 & 68\,726 & 231\,447\\
	\hline
	\(42S(n-3) + 24 \) & 6\,744 & \cellcolor{yellow} 22\,536 & 70\,584 & 222\,036 \\
	\hline
	\(127S(n-4) + 68 \) & 5\,656 & 20\,388 & 68\,140 & 213\,428\\
	\hline
	\hline
	n & 12 & 13 & 14 & 15 \\
	\hline
	\(4S(n-1) + 2 \) & \cellcolor{yellow} 815\,314 & 2\,554\,194 & 8\,045\,162 & \cellcolor{yellow} 27\,061\,154\\
	\hline
	\(13S(n-2) + 8 \) & 792\,332 & \cellcolor{yellow}2\,649\,772 & 8\,301\,132 & 26\,146\,778 \\
	\hline
	\(42S(n-3) + 24 \) & 747\,750 & 2\,559\,840 & \cellcolor{yellow} 8\,560\,800 &  25\,886\,224 \\
	\hline
	\(127S(n-4) + 68 \) & 671\,390 & 2\,261\,049 & 7\,740\,464 & 25\,886\,224 \\
	\hline
\end{tabular}
\end{center}

>>>>

Let \((p, q) \in \mathbb{N}^2\) such that there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free subsets. Let \(a \in \mathbb{N}\) with \(a > q\)
and let's try to build a coloring of \([\![1, ap + q]\!]\) into \(n + k\) weakly sum-free subsets. Let
\(l = a - b - 1\), \(r \in [\![1,q]\!]\) and \(w = a - l - r - 1 = b - r\).

@@

\noindent \underline{\textbf{\(\mathcal{T}\) block}}
\par
We color this block using the weakly sum-free coloring of \([\![1,q]\!]\) with colors \(1, ..., n\).

\noindent \underline{\textbf{\(\mathcal{R}\) block}}
\par
In this block,  we use the colors \(n + 1, ..., n + k\). We color an integer \(x\) according to its line number (written \(\lambda(x)\)).
For every \(x \in \mathcal{R}\), we color \(x\) with \(n + c\) where \(c\) is the color of \(\lambda(x)\) in the sum-free coloring of  \([\![1,p]\!]\).
Let \((x, y) \in \mathcal{R}^2\). The cases are twofold.

\begin{itemize}
	\item \underline{\(\lambda(x+y) = \lambda(x) + \lambda(y)\)} \\
	In this case, we use the sum-free property of the coloring of \([\![1,p]\!]\) (in block \(\mathcal{C}\), we only use colors \(1, ..., n\)).

	\item \underline{\(\lambda(x+y) \neq \lambda(x) + \lambda(y)\)} \\
	In this case, we do not have information about the color of \(\lambda(x+y)\). Thereby, we want to have \(x+y \in \mathcal{C}\).
	A simple solution is to limit the horizontal movement, that is if the sum changes line, not to move too far so that it stays in \(\mathcal{C}\).
	There, the maximal displacement to the left (resp. to the right) is \(2l\) (resp. \(2r\)). Not crossing entirely \(\mathcal{C}\) by going to the left
	is then expressed as \(-2l > -a + r\). Likewise, not going to far to the right is expressed as \(2r < a - l\). It can then be written as
	 \(\max(l, r) \leqslant w\).
\end{itemize}

\noindent \underline{\textbf{\(\mathcal{C}\) block}}
\par
In this block,  we use colors \(1, ..., n\). We color an integer \(x\) according to its column number, denoted by \(\tilde{\pi}(x)\). It is linked to the
projection on the first line, denoted by \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\). A simple solution is to color \(x\) with the same color
as \(\tilde{\pi}(x)\) in the weakly sum-free coloring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colors \(1, ..., n\) are sum-free.

@@

\begin{definition}
Let \( (a,n,b) \in (\mathbb{N}^*)^3\), Let \((A_1,...,A_n)\) a partition of  \([\![1, a + b]\!]\).
This partition is said to be a \(b\)-weakly-sum-free template (\(b\)-WSF-template) with width \(a\) and \(n\) colors if:

\begin{itemize}
\item \(\forall i \in [\![1, n]\!], \quad A_i\) is weakly-sum-free
\item \(\forall i \in [\![1, n]\!], \quad A_i\backslash [\![1, b]\!]\) is sum-free
\item For \(A_n\) (the special subset): \[\forall (x,y) \in A_n^2, \,
x+y>b+2a \implies x+y-2a\notin A_n\]
\item For the others subsets: \\ \[\forall i \in [\![0,n-1]\!], \, \forall(x,y) \in A_i^2, \,
x+y>a+b \implies \pi(x+y) \notin A_i
\]
\end{itemize}
\end{definition}

@@

\begin{definition}
Let \( (n,b) \in (\mathbb{N}^*)^2\). If there exists \(a\) such that there exists a \(b\)-WSF-template with width \(a\) 
and \(n\) colors, define:
\[
\WS_b^+(n)=-b+\max \{a \in \mathbb{N}^*/ \text{there exists a } b \text{-WSF-template with width } a \text{ and } n \text{ colors} \}
\]
If no such \(a\) exists, set \(\WS_b^+(n) = 0\).
\end{definition}

\begin{definition}
Let \( n \in \mathbb{N}^*\). Define 
\[
\WS^+(n)=\max_{b\in \mathbb{N}^*} \WS_b^+(n)
\]
\end{definition}

\begin{proposition}
Let \(n \in [\![2, +\infty]\!]\). Then
\[
\frac{3}{2} \WS (n-1)+1 \leqslant \WS^+(n) \leqslant \WS (n)
\]
\end{proposition}

\begin{proof}[\textsc{Proof.}]
The lower bound comes from the analogous of Abott and Hanson's construction for weak Schur numbers.
The upper bound comes from the fact that a WSF-template with width \(a\) and \(n\) colors contains a partition of
\([\![1, a]\!]\) into \(n\) sum-free subsets.
\end{proof}

@@

\begin{theorem}
Let \((a,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \([\![1,p]\!]\) 
into \(k\) sum-free subsets and a \(b\)-WSF-template \((A_1,...,A_n)\) with width \(a\) and \(n\) colors, 
then there exists a partition of \([\![1, b+p \times a]\!]\) into \((k+n-1)\) weakly sum-free subsets.
\end{theorem}

In particular, by setting \(p = S(k)\) and \(a = \WS^+(n)\) in the last theorem, the next corollary follows.

@@

\begin{remark}
In the SF-template construction for Schur numbers, the additive constant comes from the fact that the special color does
not necessarily appear right at the begining of the repeating pattern. Likewise, \(b_{max}\) can actually be replaced by \\
\[
\max_{b \in \mathbb{N}^*} \left\{\min (A_{n+1} \backslash [\![1, b ]\!]) - 1~|~ \WS_b^+(n+1) = \WS^+(n+1) \right\}
\]
\end{remark}

\begin{proof}[\textsc{Proof.}]
Let \((a,n,b) \in (\mathbb{N}^*)^3\) and \( (p,k) \in (\mathbb{N}^*)^2\). Set \(q = a + b\). \\
We denote by \(g\) the coloring associated to the \(b\)-WSF-template  and \(h\) the one associated 
to the sum-free partition of \([\![1,p]\!]\).

\[
\begin{array}{l l}
	g : [\![1, q]\!] & \longrightarrow [\![1,n]\!] \\
	h : [\![1,p]\!] & \longrightarrow [\![1,k]\!]
\end{array}
\]

@@

\begin{proposition}
Let \((b, k, a) \in \mathbb{N}^*)^3\) and let \(f\) be a coloring associated to a \(b\)-WSF-template with width \(p\) and \(k\) colors. Let
\(c \in \mathbb{N}\) (\(c = \min (A_{k+1} \backslash [\![1, b ]\!]) - 1\) works) and assume there there exists a coloring \(g\) of
\([\![b + 1, b + c]\!]\) with \(k\) colors such that:

@@

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the coloring \(i \longmapsto g(i - p S(n))\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

\par These modifications work as follows. One may notice that the first row (excluding the "tail") has constraints that other rows
do not have because of the tail, especially if the special color appears in the tail as well. Thus allowing to have a coloring on the
first row different from the coloring of the other rows would weaken the constraints. Acutally, one may even go further by
noticing that on the one hand the first (ordered) color of the sum-free partition used for the extension procedure has more
more constraints than the other colors of the sum-free partition since the first row is of this color and is more constrained than
the other rows, but that on the other hand it has more degrees of freedom than the other colors of the sum-free partition since in
the sum-free partition there cannot be two consecutive numbers of this color. As a result, it removes  some constraints imposed
 by the first row on the other rows.

\par
To sum up, one can look for a generalised WSF-template that uses a special coloring for the tail and the first row, a
coloring dedicated to the rows whose number is not 1 but is in the first color in the sum-free partition, a coloring for all
the other rows and a special coloring for the last numbers (as previously explained for the improvement of the additive
constant of WSF-templates).

@@

\begin{theorem}
Let \((k,p), (n,a) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template with width \(p\) and \(k+1\) colors and a 
\(WSF\)-template with width \(a\) and \(n\) colors, then there exists \(WSF\)-template with width \(pq\) and \((n+k)\)
 colors.
\end{theorem}

@@

\begin{corollary}
Let \(n, k \in \mathbb{N}^*\). Then
\[
\WS^+(n+k) \geqslant S^+(k+1) \WS^+(n)
\]
\end{corollary}

@@

The first two inequalities were found by Rowley, they are detailed in \cite{RowleyWS}. The third inequality is optimal and was found with a SAT solver.
It uses the first sophistication explained in the previous subsection in order to add the last number in the first color.
As for the fourth inequality, it was obtained by combining an optimal SF-template with width 33 with a WSF-template with width 4.
The best template we could get with a computer search gives the inequality \(\WS(n+4) \geqslant 127 S(n) + 68\).
It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of
5 colors which start with a good \(\WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}.
This could be the suject of a future work. Further details about the encoding as a SAT problem can be found in \cite{Heule2017}.

@@

\renewcommand{\arraystretch}{0.2}

\[
\begin{array}{c}
	\resetarraystretch
	\begin{NiceArray}{cwc{8ex}wc{10ex}wc{10ex}wc{11ex}}[hvlines]
	\CodeBefore
		\cellcolor{yellow}{2-4}
		\cellcolor{yellow}{2-5}
		\cellcolor{yellow}{3-2}
		\cellcolor{yellow}{4-3}
	\Body
		n & 8 & 9 & 10 & 11 \\
		4 \, S(n-1) + 2 & 6\,722 & 21\,146 & 71\,214 & 243\,794 \\
		13 \, S(n-2) + 8 & 6\,976 & 21\,848 & 68\,726 & 231\,447 \\
		42 \, S(n-3) + 24 & 6\,744 & 22\,536 & 70\,584 & 222\,036 \\
	\end{NiceArray}
	\\ \\
	\resetarraystretch
	\begin{NiceArray}{cwc{8ex}wc{10ex}wc{10ex}wc{11ex}}[hvlines]
	\CodeBefore
		\cellcolor{yellow}{2-2}
		\cellcolor{yellow}{2-5}
		\cellcolor{yellow}{3-3}
		\cellcolor{yellow}{4-4}
	\Body
		n & 12 & 13 & 14 & 15 \\
		4 \, S(n-1) + 2 & 815\,314 & 2\,554\,194 & 8\,045\,162 & 27\,061\,154 \\
		13 \, S(n-2) + 8 & 792\,332 & 2\,649\,772 & 8\,301\,132 & 26\,146\,778 \\
		42 \, S(n-3) + 24 & 747\,750 & 2\,559\,840 & 8\,560\,800 &  25\,886\,224 \\
	\end{NiceArray}
\end{array}
\]
